R tools for accessing research literature for text mining

### 2019-04-12 manuscript.Rmd
work on ms

---
layout: review, 11pt
linenumbers: true
title: "rOpenSci tools for accessing research literature for text mining"
author:
  - name: Scott Chamberlain
    affiliation: cstar
    email: myrmecocystus(at)gmail.com
    footnote: Corresponding author
address:
  - code: cstar
    address: |
      rOpenSci, Museum of Paleontology, University of California, Berkeley, CA, USA
abstract: |
      Corresponding Author:

      Scott Chamberlain

      rOpenSci, Museum of Paleontology, University of California, Berkeley, CA, USA

      Email address: [myrmecocystus@gmail.com](mailto:myrmecocystus@gmail.com)

      Background. xxxx.

      Methods. xxxx.

      Results. xxxx.

      Discussion. xxxx.

bibliography: components/references.bib
csl: components/peerj.csl
documentclass: components/elsarticle

output:
  pdf_document:
    template: components/elsarticle.latex
    keep_tex: true
    fig_caption: true
---

@@

There's more than 100 million articles published
(source: Crossref API), representing an enormous amount
of knowledge.  In addition to simply reading these articles,
they contain a vast trove of information of interest to
researchers for machine aided questions.

@@

Text-mining is the broad term associated with pulling information
out of articles. Given the importance of text-mining, good
text-mining tools are needed to make it easier for researchers.
In particular, the R programming language is used widely throughout
many academic fields and thus tools in R for text mining are of
particular importance.

Here, we present an overview of text-mining tools in the R programming
language. We do not cover analysis tools per se, but rather those
tools for searching for, acquiring, and "mashing up" text.

@@

Those articles that are digital can be split into two groups:
easily machine readable and non-machine readable.

The machine readable articles are those in XML, JSON, or plain text
format. The former two, XML and JSON, are ideal for the machine
readable types because they are structured data, whereas
plain text has no structure - it's simply a set of characters
with line breaks and spaces in between.

Of the non-machine readable kind, there's PDFs. These can be broken out
into two groups: text based PDFs and scanned PDFs. The former are
converted from digital versions of various kinds (MS Word, OpenOffice,
markdown, etc.), while the latter are PDFs created by scanning
in print articles for which there is no digital version.

@@

Acces to full-text is the holy grail in text-mining. Some use cases can
get by with article metadata (authors, title, etc.), some with abstracts,
but many use cases need full-text.

@@

A promising sign is an increasing number of open access publishers, yet
these represent a very small portion of the total articles (XXXXX) (ref.).

With respect to paywalled articles, access varies by institution, depending
on what each institution decided to pay for. In addition, some users may
not realize access varies with IP address so that access from campus vs.
from home (if not on a VPN) will drastically differ.

We can not end this section without mentioning SciHub. This is a last
resort option for many probably, providing access to full text of articles
that are normally paywalled. No tools in this manuscript provide access
to SciHub.

@@

xxx

#

xxx

@@

xxxxx

@@

xxxxx

@@

xxxxx

@@

xxx

@@

The following is a breakdown of the major distinct functional parts of 
`fulltext`.

@@

`ft_search()` provides search access to nine different data sources (PLOS, BMC, Crossref,
Entrez, arXiv, bioRxiv, Europe PMC, Scopus, Microsoft Academic), creating a
mostly unified interface to all data sources. The parts of each data source
that are common are mostly factored into the parameters of the `ft_search()` function,
and we also allow the user to pass on data source specific options as needed.

xxxxx

@@

`ft_links()` provides two pathways to get links (URLs) for articles, with a choice
of four different data sources (PLOS, BMC, Crossref, Entrez). First, you can
use `ft_search()`, then pass the output of that function to `ft_links()`. Second, 
you can pass DOIs directly to `ft_links()`. Both end up at the same point, links
for each article, if they could be found for the user selected data source.

@@

`ft_get()` fetchs full text of articles from many different data sources. From the DOIs
that are passed in to the function, we detect the publisher, and there are specific plugins
for certain publishers:

- aaas
- aip
- amersocclinoncol
- amersocmicrobiol
- arxiv
- biorxiv
- bmc
- copernicus
- crossref
- elife
- elsevier
- entrez
- frontiersin
- ieee
- informa
- instinvestfil
- jama
- microbiology
- peerj
- pensoft
- plos
- pnas
- royalsocchem
- sciencedirect
- scientificsocieties
- wiley

@@

@@

[rcrossref]: https://github.com/ropensci/rcrossref
[rorcid]: https://github.com/ropensci/rorcid
[fulltext]: https://github.com/ropensci/fulltext
[microdemic]: https://github.com/ropensci/microdemic
[arXiv]: https://github.com/ropensci/arXiv
[rdatacite]: https://github.com/ropensci/rdatacite
[rbhl]: https://github.com/ropensci/rbhl
[rentrez]: https://github.com/ropensci/rentrez
[rplos]: https://github.com/ropensci/rplos
[crminer]: https://github.com/ropensci/crminer
[europepmc]: https://github.com/ropensci/europepmc
[jaod]: https://github.com/ropensci/jaod
[jstor]: https://github.com/ropensci/jstor
[rcitoid]: https://github.com/ropenscilabs/rcitoid
[citecorp]: https://github.com/ropenscilabs/citecorp
[serrano]: https://github.com/sckott/serrano
[habanero]: https://github.com/sckott/habanero
[quanteda]: https://quanteda.io/
[pubchunks]: https://github.com/ropensci/pubchunks

>>>>

---
layout: review, 11pt
linenumbers: true
title: "R tools for accessing research literature for text mining"
author:
  - name: Scott Chamberlain
    affiliation: cstar
    email: myrmecocystus(at)gmail.com
    footnote: Corresponding author
address:
  - code: cstar
    address: |
      rOpenSci, Museum of Paleontology, University of California, Berkeley, CA, USA
abstract: |
  Text-mining is a powerful method for answering research questions. However,
  getting texts to extract information can be a daunting and complicated task.
  The primary reason for this is publishers. There are thousands of different
  publishers, each with their own licenses, URL patterns, access options, and
  more. Layered on top of that is the varied access each user has based on
  their institutional affiliation. Here, I introduce a suite of software packages
  in the R programming language for fetching texts. The tapestry of different
  publishers, access levels, and other factors requires a patchwork of approaches
  for getting texts to users. The flagship R package called fulltext attempts
  to simplify search and retrieval of texts for text-mining. The fulltext package,
  along with many others, make acquiring texts easier than ever, facilitating
  answering research questions with text-mining.
bibliography: components/references.bib
csl: components/peerj.csl
documentclass: components/elsarticle
output:
  pdf_document:
    template: components/elsarticle.latex
    keep_tex: true
    fig_caption: true
---

@@

There's more than 100 million research articles published
(Crossref API: <https://github.com/CrossRef/rest-api-doc>), representing
an enormous amount of knowledge.  In addition to simply reading these
articles, the articles contain a vast trove of information of interest to
researchers for machine aided questions [@Kong2018; @Usai2018].

@@

Text-mining is the broad term associated with pulling information
out of articles. Given the importance of text-mining, good
text-mining tools are needed to make it easier for researchers to do.
Graphical user interface (GUI) based text mining tools are available
[e.g., @ba2016interoperability] and some research papers have used them
[@Chaix2019], but given the urgent recent call to action for more reproducible
research [@OSC2015; @Camerer2016; @Camerer2018],
we must move away from GUI based tools as fast as possible.

In particular, the R programming language is used widely throughout many
academic fields and thus tools in R for text mining are of particular
importance because they can be adopted by academics rapidly.

Here, I present an overview of text-mining tools in the R programming
language, not for text-mining analysis, but rather those
tools for searching for, acquiring, and extracting particular chunks
of texts (e.g., title, abstract, authors). Most of the packages are
part of the rOpenSci suite (https://ropensci.org/).

@@

Those articles that are digital (which in theory includes all articles)
can be split into two groups: machine readable and non-machine
readable.

The machine readable articles are those in XML[^xml], JSON[^json], or
plain text format. The former two, XML and JSON, are the best machine
readable types because they are structured data[^sd], whereas
plain text has no structure - it's simply a set of characters
with line breaks and spaces in between.

Of the non-machine readable types, the most noteable is the Portable
Document Format (PDF)[^pdf]. These can be broken out into two groups: text based
PDFs and scanned PDFs. The former are converted from digital versions of
various kinds (MS Word, OpenOffice, LaTeX, markdown, etc.), while the
latter are created by scanning print articles to a PDF format. Text-based
PDFs are much better for text-mining purposes as plain text can be extracted
easily in R with [pdftools][], a binding to [libpoppler][]. However, with
scanned PDFs, text must be extracted using Optimal Character Recognition
(OCR; see R package [tesseract][]), which isn't always a clean solution,
especially compared to true text based PDFs.

The reality in scholarly publishing is all publishers, if they provide
any access to their articles, only provide PDF format. Very few publishers,
with some quite large (Elsevier, Pensoft, PLOS), provide XML format. Although
most publishers most likely have the XML behind each of their articles,
they for some indefensible reason do not share it - making text-mining more
difficult. Some provide plain text (Elsevier). I only know of one publisher
that provides full text as JSON (PLOS). Thus, text-mining, in most cases,
will require extracting text from PDFs.

[^xml]: https://www.w3.org/TR/xml/
[^json]: https://tools.ietf.org/html/rfc7159
[^sd]: https://en.wikipedia.org/wiki/Data_model
[^pdf]: https://en.wikipedia.org/wiki/PDF

@@

Acces to full-text is the holy grail in text-mining. Some use cases can
get by with article metadata (authors, title, etc.), some with abstracts,
but many use cases require full-text.

@@

A promising sign is an increasing number of open access articles, yet
open access articles represent a small percent of all articles: an estimate in
2018 said that 28% of the scholarly literature was open access [@Piwowar2018].

With respect to paywalled articles, access varies by institution, depending
on each institution's publisher contracts. MORE ABOUT THIS ...

Some may not realize access to articles varies with IP address so that
access from campus vs. from home (if not on a VPN) will drastically differ.
Sometimes a VPN is required, and this can provide a significant technical
hurdle to users attempting to do text-mining work.

One final hurdle in text-mining comes unsurprisingly from Elsevier. They
use so-called "fences" for programmatic access. That is, even if a person
trying to get an article programmatically their institution has access to
and they have access to, and they are on the correct IP address, they may
still not get access to an Elsevier article. Elsevier puts in place these
fences and only if you contact their technical team directly can you
get these fences removed, and only then on a per institution basis.

I can not end this section without mentioning SciHub. This is a last
resort option for many probably (or possibly first, depending on your level
of access), providing access to full text of articles that are normally
paywalled. No tools in this manuscript provide access to SciHub.

@@

A text-mining project starts with a question. From that question,
researchers then attempt to acquire scholarly articles for
text-mining. Finding those articles is not altogether straight-forward.

There are many places to search for articles; a non-exhaustive list:
Google Scholar, Microsoft Academic Research, Scopus, ScienceDirect,
Web of Science, Pubmed/Entrez, Europe PMC, Directory of Open Access Journals,
Open Knowledge Maps, and more. It's probably difficult to know where the
best place is to search. Some of these are paywalled (e.g., Web of Science),
and some are not.

The most important aspect about any source for article search with respect to
reproducible research is being able to use the data source programmatically.
Of those listed above, the following can  be used programmatically:
Microsoft Academic Research, Scopus, ScienceDirect, Pubmed/Entrez, Europe PMC,
and Directory of Open Access Journals. All of these are included in the R
package [fulltext][], discussed further below.

@@

Directory of Open Access Journals (DOAJ) maintains data on open access journals, as well
as some portion of the articles in those journals. Thus, you can search for journals as
well as articles with DOAJ. The R interface for DOAJ is [jaod][].

@@

JSTOR's Data for Research program gives institutions with access to JSTOR, access
to full text of articles within JSTOR. There is no way however to make the interaction
with JSTOR completely programmatic, thus making reproducible research very difficult.
Nonetheless, there is an R package ([jstor][]) for using data from JSTOR's Data
for Research.

@@

ORCID (https://orcid.org/) is an organization keeping track of identifiers and metadata
for researchers around the world. Individuals can optionally maintain metadata on their
scholarly works connected to their account with ORCID. Thus, across all of ORCID, a
significant cache of metadata is accruing on scholarly works, their funding amounts,
collaborators, etc., useful for bibliometrics research and more. The R interface for
ORCID is [rorcid][].

@@

The Open Citation Corpus (http://opencitations.net/) holds records of which articles
cite which other articles, allowing for all important research on the scholarly
web of citation. Citation data has been very closely guarded until recently, but the
largest publishers are still not contributing to the Open Citation Corpus. The R
interface to the Open Citation Corpus is [rcitoid][].

@@

The following is a breakdown of the major distinct parts of `fulltext`.

@@

`ft_search()` provides search access to nine different data sources (PLOS, BMC, Crossref,
Entrez, arXiv, bioRxiv, Europe PMC, Scopus, Microsoft Academic), creating a
mostly unified interface to all data sources. The parts of each data source
that are common are for the most part factored out into the parameters of the
`ft_search()` function: query term(s), pagination (number of results, result
number to start at). In addition, we allow the user to pass on data source specific
options to refine the search per data source.

With `ft_search()`, you can query any combination of the nine data sources at once.
The returned object is a list, with access to results of each data source by its name
(e.g., `$plos`, or `$crossref`). For each data source, the returned object does vary
because the returned data from each data source widely varies; for the most part
data.frame's are returned. For those data sources not queried, their slot is empty.

One important aspect of the research result we highlight is the licenses in the
returned data for each data source.

```{r}
x <- ft_search(query = 'ecology', from = c("plos", "crossref"))
```

The results for this PLOS search have all CC-BY licenses

```{r}
x$plos
```

Whereas the results for this Crossref search have mixed licenses

```{r}
x$crossref
```

You can dig into the license field for each article, with URLs holding information
on each license

```{r}
vapply(x$crossref$data$license, function(w) w$URL[1], "")
```

@@

`ft_links()` provides two pathways to get links (URLs) for articles, with a choice
of four different data sources (PLOS, BMC, Crossref, Entrez). First, you can
use `ft_search()`, then pass the output of that function to `ft_links()`.

```{r}
out <- ft_search(query = "ecology", from = "entrez")
ft_links(out)
```

Second, you can pass DOIs directly to `ft_links()`. Both end up at the same point, links
for each article, if they could be found for the user selected data source.

```{r eval=FALSE}
# FIXME
ft_links(out$entrez$data$doi)
```

@@

`ft_get()` fetchs full text of articles from many different data sources. From the DOIs
that are passed in to the function, we detect the publisher, and there are specific plugins
for certain publishers: AAAS, American Institute of Physics, American Society of
Clinical Oncology, American Society for Microbiology, arXiv, bioRxiv, BiomedCentral,
Copernicus, Crossref, Elife, Elsevier, Pubmed/PMC via NCBI's Entrez, Frontiers, IEEE,
Informa, Instituto de Investigaciones Filologicas, American Medical Association,
Microbiology Society, PeerJ, Pensoft, PLOS, PNAS, Royal Society of Chemistry,
ScienceDirect, Scientific Societies, and Wiley.

@@

To make text mining easier:

1. publishers should provide XML if they have it
2. publishers should not change URL patterns so often, or at all
3. publishers should keep their Crossref metadata up to date
4. remove all paywalls (Easy, yes?)
5. ...

@@

[rcrossref]: https://github.com/ropensci/rcrossref
[rorcid]: https://github.com/ropensci/rorcid
[fulltext]: https://github.com/ropensci/fulltext
[microdemic]: https://github.com/ropensci/microdemic
[arXiv]: https://github.com/ropensci/arXiv
[rdatacite]: https://github.com/ropensci/rdatacite
[rbhl]: https://github.com/ropensci/rbhl
[rentrez]: https://github.com/ropensci/rentrez
[rplos]: https://github.com/ropensci/rplos
[crminer]: https://github.com/ropensci/crminer
[europepmc]: https://github.com/ropensci/europepmc
[jaod]: https://github.com/ropensci/jaod
[jstor]: https://github.com/ropensci/jstor
[rcitoid]: https://github.com/ropenscilabs/rcitoid
[citecorp]: https://github.com/ropenscilabs/citecorp
[serrano]: https://github.com/sckott/serrano
[habanero]: https://github.com/sckott/habanero
[quanteda]: https://quanteda.io/
[pubchunks]: https://github.com/ropensci/pubchunks
[pdftools]: https://github.com/ropensci/pdftools
[libpoppler]: https://poppler.freedesktop.org/
[tesseract]: https://github.com/ropensci/tesseract
